Refunctionalization of Abstracting Abstract Machines (Functional Pearl)

### 2018-03-17 paper.tex
update intro

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  Abstracting abstract machines (AAM) is a systematic methodology
  for constructing abstract interpreters that are derived from concrete
  small-step abstract machines. Recent progress applies the same idea
  on definitional interpreters, and obtains big-step
  abstract definitional interpreters (ADI) written in monadic style.
  Yet, the relations between small-step abstracting abstract machines and
  big-step abstracting definitional interpreters is not well studied.

@@

Abstracting abstract machines (AAM) is a systematic methodology proposed by~\citeauthor{van2012systematic}
for constructing abstract interpreters of higher-order functional languages~\cite{van2012systematic, van2010abstracting}.
Given a concrete small-step abstract machine, (e.g, the CESK machine, Krivine's machine, etc.),
by allocating continuations in the store and bounding both the value addresses and continuation
addresses to be finite, we obtain an abstract interpreter with a finite state space which
can be used for performing sound static analysis.
One can further instantiate different polyvariant control flow analyses by using different address allocators \cite{Gilray:2016:ACP:2951913.2951936}.

Applying the same idea to big-step definitional interpreters, \citeauthor{darais2017abstracting}
built abstract definitional interpreters (ADI) that are written in monadic style \cite{darais2017abstracting}.
One of the advantages of a monadic interpreter is that it is modular and composable.
By changing the underlying monads, the definition of the interpreter is not modified,
but we can recover different semantics, including the concrete semantics and various
abstract semantics such as context-sensitivity and abstract garbage collection.

Broadly speaking, abstract abstract machines and abstract definitional interpreters are both
abstract interpreters. They are obtained by applying a combination of abstractions to their
concrete counterparts, abstract machines and definitional interpreters, respectively.
In the concrete world, the relations among reduction semantics, abstract machines,
definitional interpreters, and monadic interpreters is intensively studied by Danvy and his
collaborators in the past years
\cite{Ager:2003:FCE:888251.888254, Danvy:2001:DW:773184.773202, danvy2004refocusing, Danvy:2008:DIP:1411204.1411206, AGER2004223, ager2005functional, Danvy:2006:RW:2171265.2171268, danvy2009towards, biernacka2009towards}.
These abstract machines implement structural operational semantics in continuation style,
where the reduction contexts are defunctionalized continuations. One can derive definitional
interpreters by refunctionalizing the evaluation contexts of abstract machines,
and by defunctionalizing the higher-order functions, one may obtain abstract machines in the
reverse direction.

But in terms of their abstract semantics artifacts,
the relations between small-step abstract abstract machines and big-step abstract
definitional interpreters, as well as the ability to derive one from the other is not
well-studied.
The fundamental difference between concrete semantics artifacts and abstract semantics
artifacts is nondeterminism. In addition to ensuring termination, abstract semantics
artifacts are usually equipped with a cache of explored state that is guaranteed to reach
the least fixed-point eventually.
In this functional pearl, we practically answer the question and relate AAM and ADI
by presenting a series of syntactical transformations on the program from
small-step abstract abstract machines to big-step abstract definitional interpreters.

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  Abstracting abstract machines (AAM) is a systematic methodology
  for constructing abstract interpreters that are derived from concrete
  small-step abstract machines. Recent progress applies the same idea
  on definitional interpreters, and obtains big-step
  abstract definitional interpreters (ADI) written in monadic style.
  Yet, the relations between small-step abstracting abstract machines and
  big-step abstracting definitional interpreters are not well studied.

@@

Defining a language by building an interpreter for it can be traced to the very
early time of programming languages research \cite{landin1966next, reynolds1972definitional}.
Nowadays, even an undergraduate student in computer science is able to build toy languages
through interpreters.
But building a sound abstract interpreter remains as an esoteric and difficult task until recent years.

\citeauthor{van2012systematic} proposed Abstracting Abstract Machines (AAM) methodology which can used
for constructing abstract interpreters of higher-order functional languages from concrete abstract machines
~\cite{van2012systematic, van2010abstracting}.
Given a concrete small-step abstract machine, (e.g, the CESK machine, Krivine's machine, etc.),
by allocating continuations in the store and bounding both the value addresses and continuation
addresses to be finite, we obtain an abstract interpreter with a finite state space which
can be used for performing sound static analysis.
One can further instantiate different polyvariant control flow analyses by using different address allocators \cite{Gilray:2016:ACP:2951913.2951936}.

Applying the same idea to big-step definitional interpreters, \citeauthor{darais2017abstracting}
built abstract definitional interpreters (ADI) that are written in monadic style \cite{darais2017abstracting}.
One of the advantages of a monadic interpreter is that it is modular and composable.
By changing the underlying monads, the definition of the interpreter is not modified,
but we can recover different semantics, including the concrete semantics and various
abstract semantics such as context-sensitivity and abstract garbage collection \cite{Sergey:2013:MAI:2491956.2491979}.

Broadly speaking, abstract abstract machines and abstract definitional 
interpreters are both some kind of abstract interpreters.
They are obtained by applying a combination of abstractions to their
concrete counterparts, abstract machines and definitional interpreters, respectively.
An interesting question would be how we can interderive these abstract semantics artifacts 
from the other one.

In the concrete world, the relations among reduction semantics, abstract machines,
definitional interpreters, and monadic interpreters is intensively studied by Danvy and his
collaborators in the past years
\cite{Ager:2003:FCE:888251.888254, Danvy:2001:DW:773184.773202, danvy2004refocusing, Danvy:2008:DIP:1411204.1411206, AGER2004223, ager2005functional, Danvy:2006:RW:2171265.2171268, danvy2009towards, biernacka2009towards}.
The abstract machines implement structural operational semantics in continuation style,
where the reduction contexts are defunctionalized continuations. One can derive definitional
interpreters by refunctionalizing the evaluation contexts of abstract machines,
and by defunctionalizing the higher-order functions, one may obtain abstract machines in the
reverse direction.

On the contrary, in terms of their abstract semantics artifacts,
the relations between small-step abstract abstract machines and big-step abstract
definitional interpreters, as well as the question of deriving one from the other is not
well-studied.
The fundamental difference between concrete semantics artifacts and abstract semantics
artifacts is nondeterminism. In addition to ensuring termination, abstract semantics
artifacts are usually equipped with a cache of explored state that is guaranteed to reach
the least fixed-point eventually.
In this functional pearl, we practically answer the question and relate AAM and ADI
by presenting a series of syntactical transformations on the program from
small-step abstract abstract machines to big-step abstract definitional interpreters.
